---
title: "Add and manipulate databases and forms using R"
output: rmarkdown::html_vignette
layout: docArticle
order: 10
category: tutorials
vignette: >
  %\VignetteIndexEntry{Add and manipulate databases and forms using R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(activityinfo)
```

## Introduction

This tutorial will provide a quick introduction to how to work with forms to create and manipulate them using R. This tutorial is about the *ActivityInfo R Package* that is a client used to access to the *ActivityInfo API*. While this is an API client, names of functions and arguments are modelled more closely to web user interface to facilitate its use.

Note: in order to fully follow this tutorial you must have an ActivityInfo user account or a trial account with the permission to add a new database. Setup a free trial here: https://www.activityinfo.org/signUp

## Adding a new database

To add a new database for this tutorial use `addDatabase(label)`.

```{r eval=FALSE, include=TRUE}

newDb <- addDatabase("A new ActivityInfo tutorial database!")

```

In order to get a list of the databases that you have access to, use `getDatabases()`.

```{r eval=FALSE, include=TRUE}

dbList <- getDatabases()

# get the first database
firstDb <- dbList[[1]]

# Make the list a data.frame
dbListDf <- do.call(rbind, lapply(dbList, data.frame))

# Get the record 
newDbRecord <- dbListDf[dbListDf$databaseId == newDb$databaseId,]

```

The `getDatabaseTree(databaseId)` call provides most important information about the database.

```{r eval=FALSE, include=TRUE}
dbTree <- getDatabaseTree(databaseId = newDb$databaseId)
head(dbTree)

# One can pull out a list of the roles by their labels
roleLabels <- as.character(lapply(dbTree$roles, function(x) x$label))

# The defaults are "readonly" and "dataentry"
readOnlyRoleId <- (dbTree$roles[[which(roleLabels=="Read only")]])$id
dataEntryRoleId <- (dbTree$roles[[which(roleLabels=="Data Entry")]])$id

```

It is possible to add database users as well using `addDatabaseUser(databaseId, name, email)` and to retrieve the list of all database users using `getDatabaseUsers(databaseId)`.

```{r eval=FALSE, include=TRUE}

addDatabaseUser(
  databaseId = newDb$databaseId, 
  name = "Tutorial database user", 
  email = "activityinfo1@example.com", 
  locale = "en", 
  roleId = readOnlyRoleId
  )
addDatabaseUser(
  databaseId = newDb$databaseId, 
  name = "Tutorial database user",
  email = "activityinfo2@example.com", 
  locale = "en", 
  roleId = dataEntryRoleId
  )

# Retrieve a list of all the database users
dbUsers <- getDatabaseUsers(databaseId = newDb$databaseId)

dbUsers

```

## Forms

Adding a new form is also possible in R in three steps:
1. Define a form schema with `formSchema(databaseId, label, elements, folderId)`
2. Add new fields using `addFormField(formSchema/formId, schema, upload)`
3. Upload the newly constructed schema to your database

To create the form field, there are a number of functions available that can be found in the documentation:

- [attachmentFieldSchema](https://www.activityinfo.org/support/docs/R/reference/attachmentFieldSchema.html)
- [barcodeFieldSchema](https://www.activityinfo.org/support/docs/R/reference/barcodeFieldSchema.html)
- [calculatedFieldSchema](https://www.activityinfo.org/support/docs/R/reference/calculatedFieldSchema.html)
- [dateFieldSchema](https://www.activityinfo.org/support/docs/R/reference/dateFieldSchema.html)
- [formFieldSchema](https://www.activityinfo.org/support/docs/R/reference/formFieldSchema.html)
- [geopointFieldSchema](https://www.activityinfo.org/support/docs/R/reference/geopointFieldSchema.html)
- [monthFieldSchema](https://www.activityinfo.org/support/docs/R/reference/monthFieldSchema.html)
- [multilineFieldSchema](https://www.activityinfo.org/support/docs/R/reference/multilineFieldSchema.html)
- [multipleSelectFieldSchema](https://www.activityinfo.org/support/docs/R/reference/multipleSelectFieldSchema.html)
- [quantityFieldSchema](https://www.activityinfo.org/support/docs/R/reference/quantityFieldSchema.html)
- [referenceFieldSchema](https://www.activityinfo.org/support/docs/R/reference/referenceFieldSchema.html)
- [sectionFieldSchema](https://www.activityinfo.org/support/docs/R/reference/sectionFieldSchema.html)
- [serialNumberFieldSchema](https://www.activityinfo.org/support/docs/R/reference/serialNumberFieldSchema.html)
- [singleSelectFieldSchema](https://www.activityinfo.org/support/docs/R/reference/singleSelectFieldSchema.html)
- [subformFieldSchema](https://www.activityinfo.org/support/docs/R/reference/subformFieldSchema.html)
- [textFieldSchema](https://www.activityinfo.org/support/docs/R/reference/textFieldSchema.html)
- [userFieldSchema](https://www.activityinfo.org/support/docs/R/reference/userFieldSchema.html)
- [weekFieldSchema](https://www.activityinfo.org/support/docs/R/reference/weekFieldSchema.html)

Some examples are presented below.

```{r eval=FALSE, include=TRUE}
fmSchema <- formSchema(databaseId = newDb$databaseId, label = "ActivityInfo form generated from R")

consentField <- singleSelectFieldSchema(
  label = "Have you agreed and signed the consent statment", 
  code = "consent", 
  description = "Each respondant should be provided with the consent form.",
  options = toSelectOptions(c("Yes","No"))
)

dobField <- dateFieldSchema(label = "When were you born?", code = "dob")

# Note that it is possible to chain together

fmSchema <- fmSchema |>
  addFormField(consentField) |>
  addFormField(dobField) |>
  addFormField(textFieldSchema(label = "What is your name?", code = "respondant"))


```

It is also possible to remove a single question with `deleteFormField(formSchema/formId, label/code/id)`.

```{r eval=FALSE, include=TRUE}

fmSchema <- deleteFormField(fmSchema, label = "What is your name?")

```

After the form is completed, it can be uploaded with `updateFormSchema(schema)`.

```{r eval=FALSE, include=TRUE}

updateFormSchema(fmSchema)

```
In order to get the form or sub-form schemas that you have access to, use `getFormSchema(formId)` call, which actually retrieves "`resources/form/{id}/schema`" from the API.

```{r eval=FALSE, include=TRUE}

getFormSchema(fmSchema$id)

```

## List all forms and database resources

It is possible to access the list of all the database forms as a data.frame with the function `getDatabaseResources(databaseTree)`.

```{r eval=FALSE, include=TRUE}

# Update the database tree as new resources and users have been added.
dbTree <- getDatabaseTree(databaseId = newDb$databaseId)

# Retrieve a list of all the resources (forms, sub-forms)
dbResources <- getDatabaseResources(databaseTree = dbTree)

dbResources

```

