---
title: "Analyze and visualize ActivityInfo Data"
author: "BeDataDriven B.V."
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyze and visualize ActivityInfo Data}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

*R* is known to be a very powerful language for the analysis and visualization.
For visualizing data, there are better existing solutions that provide quite bit
of interactivity with the *ActivityInfo* data e.g. *ActivityInfo*'s built-in
visualization tools, *Power BI*, *Tableau* etc.

However, there are number of things one can only do with *R* like advanced
analysis such as prediction, statistical analysis, text mining and so on.

*R* is a batteries-included language that has many built-in calls for many
statistical analysis. Besides, it has external package system that one of the
best ones is `ggplot` that provides powerful graphics to the users.

```{r, message=FALSE}
library(activityinfo)
```

## Build a regression model

We have an example fake dataset illustrating the building
maintainance of the high schools in the Netherlands.

We first pull the data from *ActivityInfo* and we then use *R* calls to perform
analyses. The `schools` table below is the saved & cleaned version of the data
pulled via the *ActivityInfo API R* client's `queryTable()` call.

```{r}
pat <- system.file("extdata", "schools.csv", package = "activityinfo")
schools <- read.csv(pat, stringsAsFactors = FALSE)
head(schools)
```

As an advanced analysis, we try to predict the building value based on the
square meters of the building. For this analysis we use a linear regression
model. On the one column, we have a high school that it has the square meters of
the building and on the other column, we have a building value information.

```{r}
library(ggplot2)

ggplot(schools, aes(square_meters_of_school, building_value)) +
  geom_point(aes(color = as.factor(is_painted))) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_continuous(labels = scales::dollar_format(
    scale = 1 / 1e6,
    prefix = "\u20ac",
    suffix = "M"
  )) +
  scale_color_discrete(
    name = "The building painted",
    breaks = c(1, 0),
    labels = c("Yes", "No")
  ) +
  labs(
    title = "Building value vs. sqm of building",
    x = "Square meters of building",
    y = "Building value"
  ) +
  theme_minimal()
```

## Text analysis

We count the number of words in the description field, which is a multi-line
narrative field in the *ActivityInfo*, and produce a bar chart.

```{r, message=FALSE}
library(quanteda)

n_tokens <- ntoken(char_tolower(schools$situation_description), remove_punct = TRUE)
n_tokens <- as.integer(n_tokens)
df <- data.frame(
  school = schools$school_name,
  n_tokens = n_tokens,
  stringsAsFactors = FALSE
)
df
```


```{r, fig.height = 3}
ggplot(df, aes(x = reorder(school, n_tokens), y = n_tokens)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  theme_minimal()
```

For deeper information how the textual data stored 
in the *ActivityInfo* is analyzed, 
please check the [QualMiner project](https://github.com/bedatadriven/QualMiner).

