---
title: "Introduction to ActivityInfo API"
author: "BeDataDriven B.V."
date: "`r Sys.Date()`"
layout: docArticle
category: tutorials
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Introduction to ActivityInfo API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

This introductory vignette is about the *ActivityInfo API R Client* that is an *R* package to access 
to the *ActivityInfo API*. 
With this client, you can programmatically obtain and modify your data.

For the problems related to *ActivityInfo* itself, not API, please read the
[ActivityInfo Documentation](https://www.activityinfo.org/support/docs/index.html) thoroughly or contact us.

## 1. Install the R client

First, you should install *ActivityInfo API R client* in your system.

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("bedatadriven/activityinfo-R")
```

Then, put the library call at the top of your script file.

```{r}
library(activityinfo)
```

## 2. Authentication

In order to access to the *ActivityInfo API*, an authentication process has first to be done. One must setup a personal access token using the ActivityInfo dashboard. Logging in by email and password through the API is now deprecated and insecure.

```{r, eval=FALSE}
activityInfoToken(token)
```

After the authentication process has been successfully completed, a prompt will ask if the token file named *`r basename(activityinfo:::credentialsFile)`* should be written to your home directory. Only write the file to your home directory if this is secure. Don't share or publish this token file and preserve securely as it contains your personal access token.

Before start working on the API calls, do not forget to restart your *R* session to have changes applied.

## 3. Databases

In order to get the database schemas that you have access to, use`getDatabaseSchema(databaseId)` call, which actually retrieves "`/resources/database/{databaseId}`" from the API.

```{r}
db_schema <- getDatabaseSchema("ck35o0kja3")
head(db_schema)
```

## 4. Forms

In order to get the form or sub-form schemas that you have access to, use `getFormSchema(databaseId)` call, which actually retrieves "`resources/form/{id}/schema`" from the API.

```{r}
form_schema <- getFormSchema("ck35pixlz9")
head(form_schema, 4)
```

## 5. Querying

#### 5.1. Querying databases/forms in "long" format 

In the ActivityInfo API, this call requests "`resources/form/{formId}/query/rows`" (or `database` and `databaseId`) under the hood. The same call can be requested with `query/rows`; however, since *R* stores its data in column major format, there should generally be no reason to use `query/rows` from *R*.

```{r, warning=FALSE}
records <- getDatabaseRecordTable("ck35o0kja3", as.single.table = TRUE)
head(records)
```

#### 5.2. Query individual forms as data.frame

It's also possible to query individual forms as a `data.frame`. With the `columns` argument, it's possible to specify the query parameters for which columns to fetch and that can also be used to rename the column names.

```{r}
qt <- queryTable("ck2ll1o1ec", columns = c(
  id = "_id",
  sum = "Parent.sum",
  sqm = "[Square Meters]"
))
head(qt)
```

The notation used in the columns text values are based on the *ActivityInfo Query Language* and more information can be read in the [Formulas Manual](https://www.activityinfo.org/support/docs/m/77022/index.html).

